---
title: 'Absorption Spectrum of Algae'
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3)
```

## Pre-Setup

Today we will be making a graph that illustrates the absorption rates of your algae samples at different wavelengths. When you are done, your graph will look very similar to the output of the spectrophotemeter you used during the lab. However, you will be able to visualize the similarities and differences of all three algae species at once to make easy comparisons. 

## 1. Load Data Into R

We will use a similar procedure as the last lab to load data into R so that we can work with it.

* Make a new project in a new folder and open a new script.
* Load libraries you will need: `ggplot2`, `gsheet`, and  `dplyr`

For this lab we will provide you with the directions of what to do by showing you the comments you should enter in you R script.

\textbf{This lab has NO NEW code. Everything in this lab you have seen before. Therefore, code will NOT be highlighted in red}

* Copy the comment lines into your script.
* Enter the necessary code based on the previous lab. In any case where you have not seen the necessary code before, it is provided.
* Answer questions in this assignment using additional comments.

```{r load1}
# Load the libraries ggplot2, gsheet, and dplyr. 
```

```{r load2, echo=FALSE, message=FALSE}
library(ggplot2)
library(gsheet)
library(dplyr)
```

Next, load in the data. Today's data is found at:

https://docs.google.com/spreadsheets/d/1Dbo-YNgQ3ecQrz-NB8l3a_a-J2Y6laJKY7AyROF2bvc/edit?usp=sharing

```{r load3}
# Assign website address for the data to a variable
```

```{r load4, echo=FALSE}
url <- 'https://docs.google.com/spreadsheets/d/1Dbo-YNgQ3ecQrz-NB8l3a_a-J2Y6laJKY7AyROF2bvc/edit?usp=sharing'
```

```{r }
# Load the data from google sheets
```

*  Note: the following instructions show assigned the data to a variable named algae_data. You can use this name and follow these instructions or change to a new variable name.

```{r load5, echo=FALSE}
algae_data <- gsheet2tbl(url)
```

Remember, once we have our data in, it is always good to check to make sure the data was imported correctly. Use `head` to check your imported data. 
```{r head, eval=FALSE}
# Check your data looks right (first lines only)
head(algae_data)
```

* Note: here the data are called `algae_data` - your may have a different name

## 2. Graphing the Algae Data

Just like last time, we are going to use `ggplot` to graph the data. We want to compare which wavelengths are most strongly and poorely absorbed by which species of algae. We are first going to graph the data as points. 

### First create the base layer of your plot

* What is the independent variable (x axis)?
* What is the dependent variable (y axis)?
* How will you plot light and dark conditions separately?

```{r plot1a, echo=FALSE}
ggplot(algae_data, aes(x=Wavelength, y=Absorption, color=Algae))
```

### Second, add the data to the plot (using `geom_point`)
```{r plot1b, echo=FALSE, warning=FALSE}
ggplot(algae_data, aes(x=Wavelength, y=Absorption, color=Algae))+
  geom_point()
```

What wavelengths are most strongly (and most poorly) absorbed by which algae extracts?
Does your data match the known absorption spectra of different pigments found in chloroplasts?
Which pigments have absorption spectra most similar to your results? 


