---
title: "Cellular Respiration"
author: "Bio 103"
date: "Lab 4"
output: pdf_document
fontsize: 12pt 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### R Setup

* Make a new R project. If you are using a computer that does not belong to you make this project on your flash drive.

### Change in carbon dioxide concentration

* Input your carbon dioxide concentrations (ppm) at thirty-second intervals as in your lab manual (Table 4-1) using Google Sheets.
* Download the data as a tsv in your project folder.
* Load the data into R and view it. Make sure time is a variable not a label - it is the dependent variable in your analyses. Remember that in the first lab you set `row.names = 1` to make the first column a label. Do not do that in this case.
```{r, echo=FALSE, warning=FALSE}
co2 <- read.table('co2_exdata.txt',header = TRUE, sep="\t")
```

It should look like the following.
```{r, echo=FALSE, comment=NA}
co2
```

* Graph the change in carbon dioxide concentrations. Use tab-complete to make sure you get the names of the variables correct.
```{r}
plot(x = co2$Time..minutes.,y = co2$CO2.levels..glucose..ppm.)
```

* Fix your x and y axis labels

* Fit a line to the data for glucose. Note this line is a model (the lm command stands for linear model). When you fit a model to your data it is in the form y ~ x (or y as a function of x). This function allows you to estimate a value of y for a value of x.

```{r}
lm(co2$CO2.levels..glucose..ppm. ~ co2$Time..minutes.)
```

* Save your model to a variable so you can use this information later. Saving to a variable is done the same way as when you loaded data and saved it to a variable. Note the slope and the intercept. The slope is the respiration rate (ppm / min).

* Calculate the respiration rate for yeast eating lactose.

* Add the lactose data to your graph

```{r, eval=FALSE, warning=FALSE}
points(x = co2$Time..minutes.,y = co2$CO2.levels..lactose..ppm.)
```

* Add the lines from both models to the graph. a is the intercept and b is the slope.

```{r, eval=FALSE, warning=FALSE}
abline(a = -360, b = 1291)
```
* You can also plot the line from the model in one step.
```{r, eval=FALSE, warning=FALSE}
abline(lm(co2$CO2.levels..glucose..ppm. ~ co2$Time..minutes.))
```

* Change the colors of the datasets (points and lines) so they are different. You will need to replot each set of points and both lines while adding col="blue" or col="red"

* Add a legend so you know what the lines refer to. In your legend command you must include x,y coordinates, a list of the items to put in the legend, and corresponding list of the color of those items.
```{r, eval=FALSE, warning=FALSE}
legend(1,6000,c("glucose", "lactose"), text.col = c("red","blue"))
```

* Why are there no error bars on this graph?



* What can you conclude about the respiration rate of yeast?



### Calculating respiration rates

* Input the class data using Google Sheets. You should have a row for each group; each sugar is entered in a column.
* Download the data as a tsv in your project folder
* Load the data into R and view it.

```{r, echo=FALSE, warning=FALSE}
exdata <- read.table('example_data.txt',header = TRUE, sep="\t", row.names = 1)
```
It should look like the following.

```{r, echo=FALSE, comment=NA}
exdata
```

* Calculate the mean and standard deviation for the respiration rate of yeast on each sugar.

* Graph the mean respiration rate for yeast on each sugar with error bars equal to 1 standard deviation.

* Why does this graph have error bars?

* Is the respiration rate different for yeast when they eat different food? 

* Try using a boxplot as well and compare how the approaches help you understand your results.

* To save your graphs add the following prior to the lines plotting your graph
```{r, eval=FALSE, warning=FALSE}
pdf(file = 'CO2.pdf',width = 5,height = 5)
```
* Add the following after the lines plottings your graph
```{r, eval=FALSE, warning=FALSE}
dev.off()
```
* Rerun the code including these lines and your plots. This will save a pdf of your graph. You may adjust the width and height as needed.

* Save your R script as a reference for the future. Do not copy this text into Word to save it. This will add Word formatting and affect the function of your script.
